Fast and Scalable Mixed Precision Euclidean Distance Calculations Using GPU Tensor Cores

### 2025-08-28 arxiv_paper/main.tex
Created arxiv version of paper.
Stripped out comments since source will be uploaded. Changed document
class to "nonacm" to remove ACM information.

%%
%% This is file `sample-sigconf-authordraft.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,authordraft')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf-authordraft.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigconf]{acmart}
%\documentclass[sigconf,authorversion]{acmart}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{10.1145/3754598.3754636}
%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[ICPP '25]{54th International Conference on Parallel Processing}{Sept 08--11,
  2025}{San Diego, CA}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{979-8-4007-2074-1/2025/09}

@@

%Potential names for the algorithm
% Fast and Scalable Tensor Euclidean Distance
% FaSTED

% Mixed Precision Tensor Euclidean Distance
% MPTED

@@

Based on our investigation, we find that there are two major challenges that limit the performance achievable on GPU TCs for Euclidean distance calculations that we describe as follows.
\begin{enumerate}[wide=0pt]
\item \bfseries{Memory Bottlenecks --} A major challenge is efficiently transferring data between each level of the GPU's memory hierarchy, from global memory to shared memory and then into registers.

\item \bfseries{Insufficient Data Reuse --} Given the enormous disparity between FP16-32 TC throughput and global memory bandwidth, exploiting data reuse is critical. 
\end{enumerate}

@@

%We show that \ouralg can achieve up to 49\% of peak FP16-32 TC performance.

@@

{
\addtolength{\belowdisplayskip}{-1.0ex}
\addtolength{\belowdisplayshortskip}{-0.5ex}
\addtolength{\abovedisplayskip}{-2.0ex}
\addtolength{\abovedisplayshortskip}{-2.0ex}
% \begin{equation}
% dist(p_i, p_j) = \sqrt{\sum^{d}_{k=1}(p_{i,k} - p_{j,k})^2}= \sqrt{\sum^{d}_{k=1}p_{i,k}^2 - 2p_ip_j + p_{j,k}^2}.
% \label{eqn:euclidean_distance}
% \end{equation}
\begin{equation}
	\begin{aligned}
		dist(p_i, p_j) &= \sqrt{\sum^{d}_{k=1}(p_{i,k} - p_{j,k})^2} \\
									 &= \sqrt{\sum^{d}_{k=1}p_{i,k}^2 - 2p_{i,k}p_{j,k} + p_{j,k}^2}.
	\end{aligned}
\label{eqn:euclidean_distance}
\end{equation}
}

@@

\subsubsection{Scenario 2: Index-Supported Distance Similarity Searches}\label{sec:scenario2_index}
To improve the performance of brute force similarity searches (described above), algorithms will typically first construct an indexing data structure. The index is used to prune the search between points that are sufficiently far from each other, decreasing the amount of comparisons required. %While our proposed algorithm, \ouralg, does not use an index, we compare it to GPU algorithms that use an index.

@@

There are two main PTX instructions for using TCs. The first is \texttt{ldmatrix}, shown in Listing~\ref{lst:ldmatrix}, which creates a fragment of a matrix that is stored in registers shared across a warp. The second is \texttt{mma.sync}, shown in Listing~\ref{lst:mma}, which computes $\bm{D}=\bm{A}\times \bm{B}+\bm{C}$, where $\bm{A}$, $\bm{B}$, $\bm{C}$, and $\bm{D}$ are each fragments that were created by prior \texttt{ldmatrix} instructions. %The fragment $\bm{D}$ can be substituted for $\bm{C}$ to accumulate the result of the multiplication in place. 
When using FP16-32 instructions, the data stored in matrices $\bm{A}$ and $\bm{B}$ are represented in FP16 as 16$\times$16 and 16$\times$8 fragments, while the accumulator matrices, $\bm{C}$ and $\bm{D}$, are represented in FP32 as 16$\times$8 fragments.

@@

%original
% {\bf The most relevant paper in the literature regarding TC performance optimization --} Cui~\cite{cui2024acceleration} proposes using TCs for finite element methods (FEM), which are computationally expensive and thus a good target for GPU TCs. They show that data copies between shared memory and registers can be a bottleneck if there are bank conflicts. The WMMA API requires that data are accessed in row- or column-major order; therefore, solely employing the WMMA API is likely to produce unresolvable shared memory bank conflicts. Consequently, Cui~\cite{cui2024acceleration} proposes using PTX instructions that can be used to produce more sophisticated load/store patterns between shared memory and registers that eliminate bank conflicts. They achieve 45\% of the peak throughput of FP64 TCs on the Nvidia A100  GPU (roughly 8 TFLOPS). The paper leaves the problem of efficient data transfer from global to shared memory for future work.

% Similarly to Cui~\cite{cui2024acceleration}, we use PTX instructions to eliminate bank conflicts when copying data between registers and shared memory. Unlike  Cui~\cite{cui2024acceleration}, we optimize copying data from global memory into shared memory in a conflict-free manner.

@@

\section{\ouralg}\label{sec:our_algorithm}
We present our Fast and Scalable Tensor core Euclidean Distance (\ouralg) algorithm, which exploits (FP16-32) TCs on GPUs to compute Euclidean distances. We begin by describing how TCs can be used for Euclidean distance calculations and then describe the optimizations required to effectively utilize FP16-32 TCs using the A100 as our platform (the method is generalizable to other TC-equipped GPU models as well).
%\mike{Move some of this to the first MPTC-join section for motivation.}

@@

Figure~\ref{fig:single_mma} shows an illustrative example of calculating Euclidean distances, where the distances between points $p_1, \ldots,p_{16}$ and $p_1,\ldots,p_8$ are computed.  The steps below describe how TCs can be used to calculate the resulting set of 128 Euclidean distances:
%We start with all the coordinates of all points $p_1,\ldots,p_{16}$ in dataset $D$ in row-major order in global memory.

@@

% On the A100 GPU, the peak throughput of TCs operating in FP16-32 mode is 312~TFLOPS. 
In a matrix multiply-accumulate product, there are two floating-point operations for every two elements processed. Peak throughput (312~TFLOPS) requires reading the following number of elements per second:
\[
\frac{\text{312 TFLOP}}{\text{second}} \times\frac{\text{2 elements}}{\text{2 FLOP}} = \frac{\text{312}\times 10^{12}~\text{elements}}{\text{second}}.
\]

@@

{\bf (2)} Shared memory bandwidth is 17.9 TB/s. Therefore each value read from shared memory must be reused:
\[
\frac{\text{312}\times 10^{12}~\text{elements}}{\text{second}} \times \frac{\text{2 B}}{\text{1 element}} \times \frac{\text{second}}{\text{17.9 TB}} = \textbf{35 times}.
\]
   % \textbf{Conditions (1) \& (2) must be satisfied to realize peak performance.}
\end{titled-frame}

@@

% \subsection{Hierarchical Memory Reuse and Optimization}
\subsection{Optimizations: Mitigating the Memory Bottleneck}\label{sec:ouralg_optimizations}

@@

%To mitigate high latency global memory accesses, 
The L2 cache has a bandwidth of 6.4 TB/s which is $\sim$4$\times$ greater than global memory. Our algorithm is able to utilize most of this bandwidth by having blocks retrieve work from a queue such that concurrently executing block tiles read similar data elements (point coordinates) to maximize spatial locality across SMs (Figure \ref{fig:entire_mma}).

@@

%(there are 108 SMs on the A100). Thus, the total number of blocks is independent of dataset size or dimensionality, as we use 108$\times$2 $=$216 blocks.

@@

\subsubsection{Shared Memory Alignment}\label{opt:smemalignment}
Shared memory allocations do not have alignment guarantees. 
%\footnote{\url{https://docs.nvidia.com/cuda/cuda-c-programming-guide/index.html\#device-memory-accesses}} 
Alignment specifiers,\\ \texttt{\_\_align\_\_(128)}, were added to shared memory definitions to ensure 128~B alignment. This enables swizzling to place the data in the appropriate bank, thus mitigating bank conflicts.

@@

% \emph{Tensor Core Pipe Utilization:} The percentage utilization of the TCs on the GPU, as reported by the Nvidia Nsight Compute profiler.

@@

% Both \tedjoinbrute and \tedjoinindex are configured with 256 threads per block and 8 query points per warp. \tedjoinindex aborts the distance calculation when the running total distance exceeds the search distance, $\epsilon$.

@@

\begin{table}[!t]
\caption{Comparison of implementation properties.}
\footnotesize
    \centering
    {
        % Get Ted-Join-Brute to fit on one line
        % \setlength{\tabcolsep}{4pt} % default is 6pt
		\begin{tabular}{l|l|l|>{\raggedright\arraybackslash}p{1.2cm}|>{\raggedright\arraybackslash}p{1.8cm}} 
        \hline
        \rowcolor{blue!40!white}Implementation&GPU Core&Precision& Scenario~1: Brute Force  (Sec.~\ref{sec:scenario1_brute_with_epsilon})&Scenario~2: Index-Supported (Sec.~\ref{sec:scenario2_index})\\ \hline
        \ouralg&Tensor&FP16-32&\checkmark&\\
        \tedjoinbrute&Tensor&FP64&\checkmark&\\
        \tedjoinindex&Tensor&FP64&&\checkmark\\
        \gdsjoin&CUDA&FP32&&\checkmark\\
        \mistic & CUDA & FP32 & &\checkmark\\
        \hline
        \end{tabular}
    }
    \label{tab:implementations}
\end{table}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/RealDataSetSpeedComparison.pdf}
    \caption{Comparison of \ouralg to the SOTA GPU index-supported similarity search algorithms (\mistic, \gdsjoin, and \tedjoinindex) on four real-world datasets. The total end-to-end response time is plotted, which includes all associated overheads for each method (e.g., index construction and transferring data to/from the GPU). \tedjoinindex exceeds shared memory at $d{=}384$, so it is not shown in panels (c)--(d). The speedup of \ouralg over the reference implementations is reported inside each bar. The standard deviation of the response times for \ouralg in each plot is (a) $\sigma=42$, (b) $\sigma=0.11$, (c) $\sigma=0.12$, and (d) $\sigma=0.054$. In other words, the difference between time trial measurements is negligible on all datasets except for \cifar because the execution time was negligible on that dataset. }
%    \mike{I don't understand the relative standard deviation reported here, also it does not specify which algorithm. I would simply write: ``The standard deviation of the response times for \ouralg in each plot is (a) $\sigma=XXX$, (b) $\sigma=XXX$, (c) $\sigma=XXX$, and (d) $\sigma=XXX$, in other words the difference between time trial measurements is negligible on all datasets except for \cifar because the execution time was negligible on that dataset.''}
    \label{fig:realworldcomparison}
		\Description{Four plots showing how \ouralg performance compares to three other euclidean distance algorithms. Each plot depicts a different real-world dataset. Each dataset was tested with three different selectivity levels as well. }
\end{figure*}

@@

\begin{table}[!t]
\caption{ The accuracy of \ouralg compared to the FP64 \gdsjoin algorithm using the overlap accuracy metric (Equation 3) across the three selectivity levels for all real-world datasets.}
%\mike{Need to update the caption because it says three selectivity levels and we only have one selectivity for the distance error metric.}

%MG: I thought this was confusing because in a sense, we're examing error statistics.
%BC: I see, if the reviewer complains that there aren't confidence intervals, we can rebut.
%Error statistics are not included as accuracy calculations are deterministic for a given selectivity.
\footnotesize
    \centering
		\begin{tabular}{p{3.5cm}|r|r|r|r} 
        \hline
             \rowcolor{blue!40!white} & \sift & \tinyds& \cifar & \gist  \\\hline
             % Accuracy Score (Equation 3)& 1.0 & 0.99998 &  0.99971 &  0.99999 \\ 
             %\multicolumn{5}{c}{Accuracy Score (Equation~\ref{eqn:accuracy})}\\\hline 
             $S_{s}{=}64$& 1.0 & 0.99998 &  0.99971 &  0.99999 \\ 
             $S_{m}{=}128$& 1.0 & 0.99997 &  0.99955 &  0.99998 \\ 
             $S_{l}{=}256$& OOM & 0.99996 &  0.99946 &  0.99997 \\ \hline
        \end{tabular}
    \label{tab:accuracy}
              % Raw Values & 1.0 & 0.99998006551403 &  0.9997090359241025 &  0.9999852266643082 \\ \hline
\end{table}

\begin{table}[!t]
\caption{The accuracy of \ouralg compared to the FP64 \gdsjoin algorithm using the distance metric with the smallest selectivity level, $S_{s}{=}64$, for all real-world datasets.}
\footnotesize
    \centering
		\begin{tabular}{p{1.6cm}|r|r|r|r} 
        \hline
             \rowcolor{blue!40!white} & \sift & \tinyds& \cifar & \gist  \\\hline
             %\multicolumn{5}{c}{Distance error statistics for $S_{s}{=}64$}\\\hline 
             Mean & $2.6\times 10^{-6}$ & $-1.5\times 10^{-7}$ & $-5.2\times 10^{-7}$  & $-1.6\times 10^{-6}$  \\
             %Minimum & $-5.0 \times 10^{-4}$ & $-1.4\times 10^{-3}$ & $-1.6\times 10^{-4}$ & $-9.3\times 10^{-4}$ \\
             %Maximum & $5.0 \times 10^{-4}$ & $3.1\times 10^{-3}$ & $2.2\times 10^{-3}$ & $1.2\times 10^{-2}$ \\
             Std. Dev. & $2.4\times 10^{-4}$ & $9.4\times 10^{-6} $ & $3.4\times 10^{-5}$ & $3.7\times 10^{-5}$\\
             %$S_{m}{=}128$& XXX (YYY)& XXX (YYY)&  XXX (YYY)&  XXX (YYY)\\ 
             %$S_{l}{=}256$& XXX (YYY)& XXX (YYY)&  XXX (YYY)&  XXX (YYY)\\ 
             \hline
        \end{tabular}
    \label{tab:distance_accuracy}
              % Raw Values & 1.0 & 0.99998006551403 &  0.9997090359241025 &  0.9999852266643082 \\ \hline
\end{table}

@@

%\mike{Regarding the error vs. absolute error, we will need to update this text if we do the absolute error.}
%\brian{Let me just add a row for minimum error, it will show the bias in the outliers. The outliers were all biased such that we were over-estimating the distances, interestingly.}
%\brian{I am leaving it as error, and not absolute error. I think it is good to demonstrate that there is no bias and that the mean is 0. Removing the min/max seems like a good idea as it is just noise.}

@@

% Note: These statistics are computed on a subset of the total distances computed by the algorithm. Only distances that are $<\epsilon$, and appear in both the \ouralg and \gdsjoin result set are considered.

% We denote this set of intersecting points in the result set for each $p_i\in D$ as $I_a= N_i^{\ouralg} \cap N_i^{\gdsjoin}$, where $a=1,\ldots,\lvert N_i^{\ouralg} \cap N_i^{\gdsjoin}\lvert$. The distance computed between each $p_i\in D$

% \begin{equation}
% XXX
% \label{eqn:accuracy_distances}
% \end{equation}

@@

% These optimizations are able to meet the thigh demands of mixed precision tensor cores.

%There are numerous algorithms that employ Euclidean distance calculations, so our algorithm will be of great utility to many application areas outside of similarity searches.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Future work, do not delete
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%1) Add an index
%2) Upgrade FP64 Euclidean distance algorithm
%3) Model the response time of the algorithm on synthetic datasets so you know how long the algorithm will require on real-world datasets
%4) Inverstigate higher dimensional real-world datasets
%5) Investiagte lower dimensional datasets similar to those used in mistic, gds-join, etc.
%6) Apply to other problems, such as KNN, clustering, etc. 
%7) Determine if we can get 100% accuracy with FP16-32 compared to FP64 by reevaluating points that are within the margin of error of being within/outside epsilon 
%8) Run on platform with greater power budget
%9) Investigate the error in the distances computed with FP16+32 vs. FP64 and determine if we can improve the accuracy by rescaling the dataset
%10) Batching so it doesn't exceed global memory
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%End Future work, do not delete
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% XXX applications XXX

% Discussion points:
% -good for smaller $n$ larger $d$?
% \brian{Add an discussion about what to consider when running on a different GPU. }
% Future work:
% -add an index, downside here is that it'll reduce TC utilization. Would need an algorithm design that does not do index searches in the same kernel as TCs.

% \mike{From Section~\ref{sec:eval_throughput} clear future research direction is that FP64 needs to be improved.}

% \mike{Modeling response time with brute force}

%future work should probably investigate higher dimensional real-world datasts

>>>>

%%
%% This is file `sample-sigconf-authordraft.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,authordraft')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf-authordraft.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigconf,nonacm]{acmart}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
%\setcopyright{acmlicensed}
%\copyrightyear{2025}
%\acmYear{2025}
%\acmDOI{10.1145/3754598.3754636}
%%% These commands are for a PROCEEDINGS abstract or paper.
%\acmConference[ICPP '25]{54th International Conference on Parallel Processing}{Sept 08--11,
%  2025}{San Diego, CA}
%%%
%%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%%  from ``Proceedings of ...''!
%%%
%%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%%  June 03--05, 2018, Woodstock, NY}
%\acmISBN{979-8-4007-2074-1/2025/09}

@@

@@

Based on our investigation, we find that there are two major challenges that limit the performance achievable on GPU TCs for Euclidean distance calculations that we describe as follows.
\begin{enumerate}[wide=0pt]
\item \textbf{Memory Bottlenecks --} A major challenge is efficiently transferring data between each level of the GPU's memory hierarchy, from global memory to shared memory and then into registers.

\item \textbf{Insufficient Data Reuse --} Given the enormous disparity between FP16-32 TC throughput and global memory bandwidth, exploiting data reuse is critical. 
\end{enumerate}

@@

@@

{
\addtolength{\belowdisplayskip}{-1.0ex}
\addtolength{\belowdisplayshortskip}{-0.5ex}
\addtolength{\abovedisplayskip}{-2.0ex}
\addtolength{\abovedisplayshortskip}{-2.0ex}
\begin{equation}
	\begin{aligned}
		dist(p_i, p_j) &= \sqrt{\sum^{d}_{k=1}(p_{i,k} - p_{j,k})^2} \\
									 &= \sqrt{\sum^{d}_{k=1}p_{i,k}^2 - 2p_{i,k}p_{j,k} + p_{j,k}^2}.
	\end{aligned}
\label{eqn:euclidean_distance}
\end{equation}
}

@@

\subsubsection{Scenario 2: Index-Supported Distance Similarity Searches}\label{sec:scenario2_index}
To improve the performance of brute force similarity searches (described above), algorithms will typically first construct an indexing data structure. The index is used to prune the search between points that are sufficiently far from each other, decreasing the amount of comparisons required.

@@

There are two main PTX instructions for using TCs. The first is \texttt{ldmatrix}, shown in Listing~\ref{lst:ldmatrix}, which creates a fragment of a matrix that is stored in registers shared across a warp. The second is \texttt{mma.sync}, shown in Listing~\ref{lst:mma}, which computes $\bm{D}=\bm{A}\times \bm{B}+\bm{C}$, where $\bm{A}$, $\bm{B}$, $\bm{C}$, and $\bm{D}$ are each fragments that were created by prior \texttt{ldmatrix} instructions. 
When using FP16-32 instructions, the data stored in matrices $\bm{A}$ and $\bm{B}$ are represented in FP16 as 16$\times$16 and 16$\times$8 fragments, while the accumulator matrices, $\bm{C}$ and $\bm{D}$, are represented in FP32 as 16$\times$8 fragments.

@@

@@

\section{\ouralg}\label{sec:our_algorithm}
We present our Fast and Scalable Tensor core Euclidean Distance (\ouralg) algorithm, which exploits (FP16-32) TCs on GPUs to compute Euclidean distances. We begin by describing how TCs can be used for Euclidean distance calculations and then describe the optimizations required to effectively utilize FP16-32 TCs using the A100 as our platform (the method is generalizable to other TC-equipped GPU models as well).

@@

Figure~\ref{fig:single_mma} shows an illustrative example of calculating Euclidean distances, where the distances between points $p_1, \ldots,p_{16}$ and $p_1,\ldots,p_8$ are computed.  The steps below describe how TCs can be used to calculate the resulting set of 128 Euclidean distances:

@@

In a matrix multiply-accumulate product, there are two floating-point operations for every two elements processed. Peak throughput (312~TFLOPS) requires reading the following number of elements per second:
\[
\frac{\text{312 TFLOP}}{\text{second}} \times\frac{\text{2 elements}}{\text{2 FLOP}} = \frac{\text{312}\times 10^{12}~\text{elements}}{\text{second}}.
\]

@@

{\bf (2)} Shared memory bandwidth is 17.9 TB/s. Therefore each value read from shared memory must be reused:
\[
\frac{\text{312}\times 10^{12}~\text{elements}}{\text{second}} \times \frac{\text{2 B}}{\text{1 element}} \times \frac{\text{second}}{\text{17.9 TB}} = \textbf{35 times}.
\]
\end{titled-frame}

@@

\subsection{Optimizations: Mitigating the Memory Bottleneck}\label{sec:ouralg_optimizations}

@@

The L2 cache has a bandwidth of 6.4 TB/s which is $\sim$4$\times$ greater than global memory. Our algorithm is able to utilize most of this bandwidth by having blocks retrieve work from a queue such that concurrently executing block tiles read similar data elements (point coordinates) to maximize spatial locality across SMs (Figure \ref{fig:entire_mma}).

@@

@@

\subsubsection{Shared Memory Alignment}\label{opt:smemalignment}
Shared memory allocations do not have alignment guarantees. 
Alignment specifiers, \texttt{\_\_align\_\_\-(128)}, were added to shared memory definitions to ensure 128~B alignment. This enables swizzling to place the data in the appropriate bank, thus mitigating bank conflicts.

@@

@@

@@

\begin{table}[!t]
\caption{Comparison of implementation properties.}
\footnotesize
    \centering
    {
		\begin{tabular}{l|l|l|>{\raggedright\arraybackslash}p{1.2cm}|>{\raggedright\arraybackslash}p{1.8cm}} 
        \hline
        \rowcolor{blue!40!white}Implementation&GPU Core&Precision& Scenario~1: Brute Force  (Sec.~\ref{sec:scenario1_brute_with_epsilon})&Scenario~2: Index-Supported (Sec.~\ref{sec:scenario2_index})\\ \hline
        \ouralg&Tensor&FP16-32&\checkmark&\\
        \tedjoinbrute&Tensor&FP64&\checkmark&\\
        \tedjoinindex&Tensor&FP64&&\checkmark\\
        \gdsjoin&CUDA&FP32&&\checkmark\\
        \mistic & CUDA & FP32 & &\checkmark\\
        \hline
        \end{tabular}
    }
    \label{tab:implementations}
\end{table}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/RealDataSetSpeedComparison.pdf}
    \caption{Comparison of \ouralg to the SOTA GPU index-supported similarity search algorithms (\mistic, \gdsjoin, and \tedjoinindex) on four real-world datasets. The total end-to-end response time is plotted, which includes all associated overheads for each method (e.g., index construction and transferring data to/from the GPU). \tedjoinindex exceeds shared memory at $d{=}384$, so it is not shown in panels (c)--(d). The speedup of \ouralg over the reference implementations is reported inside each bar. The standard deviation of the response times for \ouralg in each plot is (a) $\sigma=42$, (b) $\sigma=0.11$, (c) $\sigma=0.12$, and (d) $\sigma=0.054$. In other words, the difference between time trial measurements is negligible on all datasets except for \cifar because the execution time was negligible on that dataset. }
    \label{fig:realworldcomparison}
		\Description{Four plots showing how \ouralg performance compares to three other euclidean distance algorithms. Each plot depicts a different real-world dataset. Each dataset was tested with three different selectivity levels as well. }
\end{figure*}

@@

\begin{table}[!t]
\caption{ The accuracy of \ouralg compared to the FP64 \gdsjoin algorithm using the overlap accuracy metric (Equation 3) across the three selectivity levels for all real-world datasets.}
\footnotesize
    \centering
		\begin{tabular}{p{3.5cm}|r|r|r|r} 
        \hline
             \rowcolor{blue!40!white} & \sift & \tinyds& \cifar & \gist  \\\hline
             $S_{s}{=}64$& 1.0 & 0.99998 &  0.99971 &  0.99999 \\ 
             $S_{m}{=}128$& 1.0 & 0.99997 &  0.99955 &  0.99998 \\ 
             $S_{l}{=}256$& OOM & 0.99996 &  0.99946 &  0.99997 \\ \hline
        \end{tabular}
    \label{tab:accuracy}
\end{table}

\begin{table}[!t]
\caption{The accuracy of \ouralg compared to the FP64 \gdsjoin algorithm using the distance metric with the smallest selectivity level, $S_{s}{=}64$, for all real-world datasets.}
\footnotesize
    \centering
		\begin{tabular}{p{1.6cm}|r|r|r|r} 
        \hline
             \rowcolor{blue!40!white} & \sift & \tinyds& \cifar & \gist  \\\hline
             Mean & $2.6\times 10^{-6}$ & $-1.5\times 10^{-7}$ & $-5.2\times 10^{-7}$  & $-1.6\times 10^{-6}$  \\
             Std. Dev. & $2.4\times 10^{-4}$ & $9.4\times 10^{-6} $ & $3.4\times 10^{-5}$ & $3.7\times 10^{-5}$\\
             \hline
        \end{tabular}
    \label{tab:distance_accuracy}
\end{table}

@@

@@

@@

@@

### 2025-08-28 arxiv_paper/tikz_src/Single_TC_MMA.tex
Created arxiv version of paper.
Stripped out comments since source will be uploaded. Changed document
class to "nonacm" to remove ACM information.

\documentclass[tikz,border=2mm]{standalone}

@@

		% Draw B
	\pgfmathsetmacro\bdXShift{\fn / 2}
	\pgfmathsetmacro\bdYShift{\fm / 2}
	\fill[style=fragment,fill=BFragmentColor] (0, 0) rectangle +(\fn, -\fk) node[style=matrixLabelStyle] {\qrf};
	\matrix (m) [style=matrixTextStyle, xshift=\bdXShift cm, yshift=-\bdYShift cm, 
	matrix of math nodes,
	row sep=1.5 cm, column sep=0.0 cm,
	nodes={anchor=center}] {
		p_{1,1} & p_{8,1} \\
		p_{1,16} & p_{8,16} \\
	};
    \draw[style=lineStyle] (m-1-1.south) -- (m-2-1.north);
    \draw[style=lineStyle] (m-1-2.south) -- (m-2-2.north);
    %\node[style=matrixLabelStyle] at (0 cm, 0 cm) {\qrf};

		% Draw D
	\pgfmathsetmacro\dXShift{0}
	\pgfmathsetmacro\dYShift{-\whiteSpace - \fk}
	\fill[style=fragment,fill=DFragmentColor, xshift=0 cm, yshift=\dYShift cm] (0, 0) rectangle +(\fn, -\fm) node[style=matrixLabelStyle] {\arf};
	\pgfmathsetmacro\adYShift{\dYShift - 0.5 * \fm}
	\matrix (m) [style=matrixTextStyle, xshift=\bdXShift cm, yshift=\adYShift cm, 
	matrix of math nodes,
	row sep=1.5 cm, column sep=0 cm,
	nodes={anchor=center}] {
		a_{1,1} & a_{1,8} \\
		a_{16,1} & a_{16,8} \\
	};
    % Draw all interconnecting lines
    %\draw (m-1-1.east) -- (m-1-2.west);
    %\draw (m-2-1.east) -- (m-2-2.west);
    %\draw (m-1-1.south) -- (m-2-1.north);
    %\draw (m-1-2.south) -- (m-2-2.north);
    \pgfmathsetmacro\matrixLabelYOffset{-\fm - \whiteSpace}
    ;

@@

	\fill[style=fragment,fill=AFragmentColor, xshift=\aXShift cm, yshift=\aYShift cm] (0, 0) rectangle +(\fk, -\fm) node[style=matrixLabelStyle] {\prf};
	\matrix (m) [style=matrixTextStyle, xshift=\matXShift cm, yshift=\adYShift cm, 
	matrix of math nodes,
	row sep=1.5 cm, column sep=1.0 cm,
	nodes={anchor=center}] {
		p_{1,1} & p_{1,16} \\
		p_{16,1} & p_{16,16} \\
	};
    \draw[style=lineStyle] (m-1-1.east) -- (m-1-2.west);
    \draw[style=lineStyle] (m-2-1.east) -- (m-2-2.west);
    \pgfmathsetmacro\matrixLabelXOffset{-\fk - \whiteSpace}
    %\node[style=matrixLabelStyle] at (\matrixLabelXOffset cm, \matrixLabelYOffset cm) {\prf};

>>>>

\documentclass[tikz, class=extarticle, 9pt]{standalone}

@@

		% Draw B
	\pgfmathsetmacro\bdXShift{\fn / 2}
	\pgfmathsetmacro\bdYShift{\fm / 2}
	\fill[style=fragment,fill=BFragmentColor] (0, 0) rectangle +(\fn, -\fk) node[style=matrixLabelStyle] {$Q_{\regf}$};
	\matrix (m) [style=matrixTextStyle, xshift=\bdXShift cm, yshift=-\bdYShift cm, 
	matrix of math nodes,
	row sep=1.5 cm, column sep=0.0 cm,
	nodes={anchor=center}] {
		p_{1,1} & p_{8,1} \\
		p_{1,16} & p_{8,16} \\
	};
    \draw[style=lineStyle] (m-1-1.south) -- (m-2-1.north);
    \draw[style=lineStyle] (m-1-2.south) -- (m-2-2.north);
    %\node[style=matrixLabelStyle] at (0 cm, 0 cm) {\qrf};

		% Draw D
	\pgfmathsetmacro\dXShift{0}
	\pgfmathsetmacro\dYShift{-\whiteSpace - \fk}
	\fill[style=fragment,fill=DFragmentColor, xshift=0 cm, yshift=\dYShift cm] (0, 0) rectangle +(\fn, -\fm) node[style=matrixLabelStyle] {$A_{\regf}$};
	\pgfmathsetmacro\adYShift{\dYShift - 0.5 * \fm}
	\matrix (m) [style=matrixTextStyle, xshift=\bdXShift cm, yshift=\adYShift cm, 
	matrix of math nodes,
	row sep=1.5 cm, column sep=0 cm,
	nodes={anchor=center}] {
		a_{1,1} & a_{1,8} \\
		a_{16,1} & a_{16,8} \\
	};
    % Draw all interconnecting lines
    %\draw (m-1-1.east) -- (m-1-2.west);
    %\draw (m-2-1.east) -- (m-2-2.west);
    %\draw (m-1-1.south) -- (m-2-1.north);
    %\draw (m-1-2.south) -- (m-2-2.north);
    \pgfmathsetmacro\matrixLabelYOffset{-\fm - \whiteSpace}
    ;

@@

\fill[style=fragment,fill=AFragmentColor, xshift=\aXShift cm, yshift=\aYShift cm] (0, 0) rectangle +(\fk, -\fm) node[style=matrixLabelStyle] {$P_{\regf}$};
	\matrix (m) [style=matrixTextStyle, xshift=\matXShift cm, yshift=\adYShift cm, 
	matrix of math nodes,
	row sep=1.5 cm, column sep=1.0 cm,
	nodes={anchor=center}] {
		p_{1,1} & p_{1,16} \\
		p_{16,1} & p_{16,16} \\
	};
    \draw[style=lineStyle] (m-1-1.east) -- (m-1-2.west);
    \draw[style=lineStyle] (m-2-1.east) -- (m-2-2.west);
    \pgfmathsetmacro\matrixLabelXOffset{-\fk - \whiteSpace}
    %\node[style=matrixLabelStyle] at (\matrixLabelXOffset cm, \matrixLabelYOffset cm) {\prf};
